Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-27 manuscript/paper.tex
polish data model & architecture

\subsection{Data model} \label{subsec:Data_Model} 
\texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of 
ARGS~\citep{wong2024general}. This efficient ARG encoding 
is implemented by the 
\texttt{tskit} library~\citep{ralph2020efficiently}
and supported by most modern ARG
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
haller2019tree,baumdicker2022efficient,
adrion2020community,lauterbur2023expanding,korfmann2023weak,
tsambos2023link,tagami2024tstrait},
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general},
and processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}.
In \texttt{tskit}, ARGs are encoded as a
collection of tables, storing information about a specific ARG
property (Fig.~\ref{fig:Figure_1}).
% For example, the Node table describes sampled or ancestral sequences
% and records their time of birth, and the individual and population they belong
% to. The Edge table describes the relationship between pairs of nodes, and
% records the node IDs and the genomic locations over which the edge spans. The
% Mutation table records mutation events in the history of the sample, including
% the change of allelic state, time of event and node in which the event
% occurred. 
The Node and Edge tables describe the genealogical relationships
between the sampled and ancestral sequences, 
while the Mutation table describes
sequence variation contained in the samples. 
\texttt{Tsbrowse} augments these tables with additional columns to create
a compressed \texttt{.tsbrowse} file, containing all necessary semantic
metadata required for automatic visualisation.
% (see \hyperref[subsec:User_Interface]{User Interface}).

\subsection{Architecture} 
\texttt{Tsbrowse} is a modular client-server application 
written in Python, optimised for handling ARGs with millions
of samples (Fig.~\ref{fig:Figure_1}). 
To ensure that such large ARGs can be visualised quickly
we perform a preprocessing step (\texttt{tsbrowse preprocess})
which augments the \texttt{tskit} 
data model with some additional derived information (e.g., the number 
of inheritors of each mutation).
To visualise an ARG we then run \texttt{tsbrowse serve}, which 
by default will open a browser window on the local machine, but 
also supports running the server on a remote machine across the 
network.
This client-server architecture has some important advantages over
a monolithic single-machine approach. Most importantly,
the ARG being visualised remains in-situ and does not need to be 
downloaded from the server.

The express goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges 
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale 
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider 
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web-browser
for display, and dynamically updates as the user interatively navigates the ARG.

>>>>

\subsection{Data model} \label{subsec:Data_Model} 
\texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of 
ARGS~\citep{wong2024general}. This efficient ARG encoding 
is implemented by the 
\texttt{tskit} library~\citep{ralph2020efficiently}
and supported by most modern ARG
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
haller2019tree,baumdicker2022efficient,
adrion2020community,lauterbur2023expanding,korfmann2023weak,
tsambos2023link,tagami2024tstrait},
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general},
and processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}.
In \texttt{tskit}, ARGs are encoded as a
collection of tables, storing information about the 
nodes and edges that describe the graph topology and the sites 
and mutations encoding the sequence variation (Fig.~\ref{fig:Figure_1}).
% For example, the Node table describes sampled or ancestral sequences
% and records their time of birth, and the individual and population they belong
% to. The Edge table describes the relationship between pairs of nodes, and
% records the node IDs and the genomic locations over which the edge spans. The
% Mutation table records mutation events in the history of the sample, including
% the change of allelic state, time of event and node in which the event
% occurred. 
% The Node and Edge tables describe the genealogical relationships
% between the sampled and ancestral sequences, 
% while the Mutation table describes
% sequence variation contained in the samples. 
% (see \hyperref[subsec:User_Interface]{User Interface}).

\subsection{Architecture} 
\texttt{Tsbrowse} is a modular client-server application 
written in Python, optimised for ARGs with millions
of samples (Fig.~\ref{fig:Figure_1}). 
The application has two basic commands: \texttt{preprocess}
and \texttt{serve}. The \texttt{preprocess} command 
takes an input ARG file and augments the \texttt{tskit} 
tables with additional columns,
precomputing all the information required for visualisation,
and storing as a compressed \texttt{.tsbrowse} file.
(The \texttt{.tsbrowse} file is also a valid input for 
\texttt{tszip}, a general utility for compressing \texttt{tskit}
files.)
To visualise an ARG we then run \texttt{tsbrowse serve}, which 
by default will open a browser window on the local machine, but 
also supports running the server on a remote machine across the 
network.
This client-server architecture has some important advantages over
a monolithic single-machine approach. Most importantly,
the ARG being visualised remains in-situ and does not need to be 
downloaded from the server.

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges 
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale 
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider 
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web-browser
for display, and dynamically updates as the user interatively navigates the ARG.
This approach allows us to summarise very large ARGs interactively;
Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of 
\texttt{tsbrowse} summarising almost 1 million mutations along the SARS-Cov-2
genome.

% Fig~\ref{fig:Figure_2} shows an example of how edges are visualised 
% under this
